Linearizability Proofs Based On Forward Simulations

### 2017-01-21 draftCAV/queue.tex
started explaining commit points
finished stacks except TS proof and completed queue stuff

\begin{wrapfigure}{l}{7cm}
\vspace{-8mm}
\includegraphics[width=7cm]{fig-queue12.pdf}
%
%\vspace{2mm}
%\includegraphics[width=7cm]{fig-queue2.pdf}
\vspace{-5mm}
\caption{Simulating queue histories with $AbsQ$.}
\label{fig:queueSim}
\vspace{-10mm}
\end{wrapfigure}
Figure~\ref{fig:queueSim} pictures two executions of $AbsQ$ for two extended histories (that include dequeue linearization points). The state of $AbsQ$ after each action is pictured as a graph below the action. The nodes of this graph represent enqueue operations and the edges happens-before constraints. Each node is labeled by a value (the input of the enqueue) and a flag {\tt PEND} or {\tt COMP} showing whether the operation is pending or completed. For instance, in the case of the first history, the dequeue linearization point $lin(deq,y,3)$ is enabled because the current happens-before contains a \emph{minimal} enqueue operation with input $y$. Note that a linearization point $lin(deq,x,3)$ is also enabled at this state.

Formally, the states of $AbsQ$ are tuples $\tup{O,<,\ell,rv,cp}$ where $O\subseteq \<Ops>$ represents a set of operation identifiers (of previously invoked enqueue operations), $<\subseteq O\times O$ is a strict partial order, $\ell: O -> \<Vals>\times\{\tt{PEND,\tt{COMP}}\}$ labels every identifier with a value and a flag that records whether the operation is pending or completed (this flag is used to track the happens-before order), $rv:\<Ops> ~> \<Vals>$ records the return value of a pending dequeue fixed at its linearization point ($~>$ denotes a partial function), and $cp:\<Ops> ~> \<Nats>$ records the control point of every pending (enqueue or dequeue) operation.
The initial state has all these components set to $\emptyset$ and the transition relation $->$ is defined in Figure~\ref{fig:transitions:AbsQ}. The alphabet of $AbsQ$ contains call/return actions and dequeue linearization points. The latter are denoted by $lin(deq,d,k)$ where $d\in \<Vals>$ and $k\in\<Ops>$. $Lin(deq)$ is the set of all such actions.

@@

\begin{figure} [t]
{\scriptsize
  \centering
  \begin{mathpar}
    \inferrule[call-enq]{
      k\not\in dom(cp) \\ 
      d\neq {\tt EMPTY}
    }{
      O,<,\ell,rv,cp
      \xrightarrow{inv(enq,d,k)}
      %O\cup\{k\},<\cup \{(k',k): \ell_2(k')={\tt COMP}\},\ell[k\mapsto (d,{\tt PEND})],rv,cp[k\mapsto 1]
      O\cup\{k\},<\cup\ {\tt COMP}(O)\times\{k\},\ell[k\mapsto (d,{\tt PEND})],rv,cp[k\mapsto 1]
    }\hspace{5mm}

    \inferrule[call-deq]{
      k\not\in dom(cp) \\ 
    }{
      O,<,\ell,rv,cp
      \xrightarrow{inv(deq,k)}
      O,<,\ell,rv,cp[k\mapsto 1]
    }\hspace{5mm}
    \inferrule[ret-deq]{
       cp(k) = 2 \\
       rv(k)=d  
    }{
      O,<,\ell,rv,cp
      \xrightarrow{ret(deq,d,k)}
      O,<,\ell,rv,cp[k\mapsto 3]
    }\hspace{5mm}

    \inferrule[ret-enq1]{
      cp(k) = 1 \\
      k \in O \\
      \ell(k) = (d,{\tt PEND}) 
    }{
      O,<,\ell,rv,cp
      \xrightarrow{ret(enq,k)}
      O,<,\ell[k\mapsto (d,{\tt COMP})],rv,cp[k\mapsto 2]
    }\hspace{5mm}
    \inferrule[ret-enq2]{
      cp(k) = 1 \\
      k \not\in O 
    }{
      O,<,\ell,rv,cp
      \xrightarrow{ret(enq,k)}
      O,<,\ell,rv,cp[k\mapsto 2]
    }\hspace{5mm}

    \inferrule[lin-deq1]{
       cp(k) = 1 \\
       d\neq{\tt EMPTY} \\
       k'\in min(O) \land \ell_1(k')=d
    }{
      O,<,\ell,rv,cp
      \xrightarrow{lin(deq,d,k)}
      O\setminus \{k'\},<\uparrow k',\ell,rv[k\mapsto d],cp[k\mapsto 2]
    }\hspace{5mm}

    \inferrule[lin-deq2]{
       cp(k) = 1 \\
       \forall o\in O. \ell_2(o)={\tt EMPTY}
    }{
      O,<,\ell,rv,cp
      \xrightarrow{lin(deq,{\tt EMPTY},k)}
      O,<,\ell,rv[k\mapsto {\tt EMPTY}],cp[k\mapsto 2]
    }\hspace{5mm}

      \end{mathpar}
  }
 \vspace{-5mm}
  \caption{The transition relation of $AbsQ$. We use the following notations: $\ell_i(k)$ denotes the projection of $\ell(k)$ over the $i$-th component, for each $i\in\{1,2\}$, ${\tt COMP}(O)=\{k\in O: \ell_2(k)={\tt COMP}\}$, $\mathit{f}[x\mapsto y]$ is the function $g$ such that $g(z)=f(z)$ for all $z\neq x$ in the domain of $f$, and $g(x)=y$, $min(O)$ is the set of elements of $O$ which are minimal in the order relation $<$, and $<\uparrow k$ denotes the relation $<$ where all the pairs containing $k$ have been removed.
  %\textcolor{red}{Call-Enq must have $d!= \texttt{EMPTY}$ as a premise. Also lin deq returning empty must be changed as before.}
  }
  \label{fig:transitions:AbsQ}
\vspace{-6mm}
\end{figure}

@@

TODO LINEARIZATION POINT

>>>>

\begin{wrapfigure}{l}{7cm}
\vspace{-8mm}
\includegraphics[width=7cm]{fig-queue12.pdf}
%
%\vspace{2mm}
%\includegraphics[width=7cm]{fig-queue2.pdf}
\vspace{-5mm}
\caption{Simulating queue histories with $AbsQ$. The order between actions is from left to right.}
\label{fig:queueSim}
\vspace{-10mm}
\end{wrapfigure}
Figure~\ref{fig:queueSim} pictures two executions of $AbsQ$ for two extended histories (that include dequeue linearization points). The state of $AbsQ$ after each action is pictured as a graph below the action. The nodes of this graph represent enqueue operations and the edges happens-before constraints. Each node is labeled by a value (the input of the enqueue) and a flag {\tt PEND} or {\tt COMP} showing whether the operation is pending or completed. For instance, in the case of the first history, the dequeue linearization point $lin(deq,y,3)$ is enabled because the current happens-before contains a \emph{minimal} enqueue operation with input $y$. Note that a linearization point $lin(deq,x,3)$ is also enabled at this state.

Formally, the states of $AbsQ$ are tuples $\tup{O,<,\ell,rv,cp}$ where $O\subseteq \<Ops>$ represents a set of operation identifiers (of previously invoked enqueue operations), $<\subseteq O\times O$ is a strict partial order, $\ell: O -> \<Vals>\times\{\tt{PEND,\tt{COMP}}\}$ labels every identifier with a value and a flag that records whether the operation is pending or completed (this flag is used to track the happens-before order), $rv:\<Ops> ~> \<Vals>$ records the return value of a pending dequeue fixed at its linearization point ($~>$ denotes a partial function), and $cp:\<Ops> ~> \{A_1,A_2,R_1,R_2,R_3\}$ records the control point of every enqueue ($A_1, A_2$) or dequeue ($R_1,R_2,R_3$) operation.
The initial state has all these components set to $\emptyset$ and the transition relation $->$ is defined in Fig.~\ref{fig:transitions:AbsQ}. The alphabet of $AbsQ$ contains call/return actions and dequeue linearization points. The latter are denoted by $lin(deq,d,k)$ where $d\in \<Vals>$, $k\in\<Ops>$. $Lin(deq)$ is the set of all such actions.

@@

\begin{figure} [t]
{\scriptsize
  \centering
  \begin{mathpar}
    \inferrule[call-enq]{
      k\not\in dom(cp) \\ 
      d\neq {\tt EMPTY}
    }{
      O,<,\ell,rv,cp
      \xrightarrow{inv(enq,d,k)}
      %O\cup\{k\},<\cup \{(k',k): \ell_2(k')={\tt COMP}\},\ell[k\mapsto (d,{\tt PEND})],rv,cp[k\mapsto 1]
      O\cup\{k\},<\cup\ {\tt COMP}(O)\times\{k\},\ell[k\mapsto (d,{\tt PEND})],rv,cp[k\mapsto A_1]
    }\hspace{5mm}

    \inferrule[call-deq]{
      k\not\in dom(cp) \\ 
    }{
      O,<,\ell,rv,cp
      \xrightarrow{inv(deq,k)}
      O,<,\ell,rv,cp[k\mapsto R_1]
    }\hspace{5mm}
    \inferrule[ret-deq]{
       cp(k) = R_2 \\
       rv(k)=d  
    }{
      O,<,\ell,rv,cp
      \xrightarrow{ret(deq,d,k)}
      O,<,\ell,rv,cp[k\mapsto R_3]
    }\hspace{5mm}

    \inferrule[ret-enq1]{
      cp(k) = A_1 \\
      k \in O \\
      \ell(k) = (d,{\tt PEND}) 
    }{
      O,<,\ell,rv,cp
      \xrightarrow{ret(enq,k)}
      O,<,\ell[k\mapsto (d,{\tt COMP})],rv,cp[k\mapsto A_2]
    }\hspace{5mm}
    \inferrule[ret-enq2]{
      cp(k) = A_1 \\
      k \not\in O 
    }{
      O,<,\ell,rv,cp
      \xrightarrow{ret(enq,k)}
      O,<,\ell,rv,cp[k\mapsto A_2]
    }\hspace{5mm}

    \inferrule[lin-deq1]{
       cp(k) = R_1 \\
       d\neq{\tt EMPTY} \\
       k'\in min(O) \\ 
       \ell_1(k')=d
    }{
      O,<,\ell,rv,cp
      \xrightarrow{lin(deq,d,k)}
      O\setminus \{k'\},<\uparrow k',\ell,rv[k\mapsto d],cp[k\mapsto R_2]
    }\hspace{5mm}
    \inferrule[lin-deq2]{
       cp(k) = R_1 \\
       \forall o\in O. \ell_2(o)={\tt EMPTY}
    }{
      O,<,\ell,rv,cp
      \xrightarrow{lin(deq,{\tt EMPTY},k)}
      O,<,\ell,rv[k\mapsto {\tt EMPTY}],cp[k\mapsto R_2]
    }\hspace{5mm}    
      \end{mathpar}
  }
 \vspace{-5mm}
  \caption{The transition relation of $AbsQ$. We use the following notations: $\ell_i(k)$ denotes the projection of $\ell(k)$ over the $i$-th component, for each $i\in\{1,2\}$, ${\tt COMP}(O)=\{k\in O: \ell_2(k)={\tt COMP}\}$, $\mathit{f}[x\mapsto y]$ is the function $g$ such that $g(z)=f(z)$ for all $z\neq x$ in the domain of $f$, and $g(x)=y$, $min(O)$ is the set of elements of $O$ which are minimal in the order relation $<$, and $<\uparrow k$ denotes the relation $<$ where all the pairs containing $k$ have been removed.
  %\textcolor{red}{Call-Enq must have $d!= \texttt{EMPTY}$ as a premise. Also lin deq returning empty must be changed as before.}
  }
  \label{fig:transitions:AbsQ}
\vspace{-6mm}
\end{figure}

@@

Finally, we show that the linearization point of a dequeue $k$ of $\mathit{HWQ}$, i.e., an execution of {\tt swap} returning a non-null value $d$, from state $s$ and leading to a state $s'$ is simulated by a transition labeled by $lin(deq,d,k)$ of $AbsQ$ from state $t$. By the definition of $\mathit{HWQ}$, there exists a unique enqueue $k_e$ which filled the position updated by $k$, i.e., ${\tt i}_s(k_e)=i_s(k)$ and ${\tt x}_{s'}(k)={\tt x}_s(k_e)$. We show that $k_e$ is minimal in the order relation $<$ of $t$ which implies that $lin(deq,d,k)$ is enabled in $t$. Thus, instantiating item (c) in the definition of $<$ with $k'=k_e$ and $k_d=k$ we get that every enqueue that reserved a position smaller than the one of $k_e$ can't be ordered before $k_e$ in the order $<$. Also, applying item (b) with $k=k_e$ we get the same for every enqueue that reserved a bigger position. An enqueue that didn't reserved a position is by definition maximal in the order $<$ and therefore, not a predecessor of $k_e$. The state $t'$ obtained from $t$ through a $lin(deq,d,k)$ transition is related to $s'$ because essentially, (1) the value added by $k_e$ is not anymore present in {\tt items} which implies that $k_e$ doesn't occur in any $AbsQ$ state related to $s'$, and (2) the value of ${\tt x}(k)$ is set to $d\neq {\tt null}$ which implies that $rv(k)$ is set to $d$ in every $AbsQ$ state related to $s'$.
